---
title: "Dropbead tutorial"
author: "Nikos Karaiskos"
date: "07 Feb 2017"
output: html_document
---

## Knee plot and the number of cells
Dropbead provides a function to computationally estimate the number of cells
present in the sample, by calculating the inflection point of the cumulative
fraction of reads against the cell barcodes. Assuming that `reads.by.cell` 
is the `data.frame` with the reads per cell (the output of `BAMTagHistogram`
from the Drop-seq toolkit), 

```{r, echo=FALSE, include=FALSE}
library(dropbead)
library(data.table)
library(xlsx)
library(rJava)
library(xlsxjars)
reads.by.cell <- data.frame(fread('zcat < /mydaten/projects/hek3t3/data/ds_013_50fix/out_readcounts.txt.gz'))
dge.matrix=data.frame(fread("zcat < /mydaten/projects/hek3t3/data/ds_013_50fix/dge.txt.gz"), row.names = 1)
```

```{r}
plotCumulativeFractionOfReads(reads.by.cell, 
                              cutoff = 10000, 
                              draw.infl.point = TRUE)
```


# Mixed species experiments
Dropbead offers two main classes to store samples, the `SingleSpeciesSample` 
class and the `MixedSpeciesSample` class, the latter containing the former.

Mixed species experiments are important to estimate the number of doublets in 
the samples. Dropbead assumes that a digital gene expression matrix (DGE)
has already been generated and the genes for the two species are separated
by a prefix. For mixed human/mouse samples the prefixes are `hg_` and `mm_` 
respectively. Assuming that the DGE has been loaded in `dge.matrix`,

```{r}
# The object containing the sample
mms <- new("MixedSpeciesSample", 
           species1="human", 
           species2="mouse",
           dge=dge.matrix)
```

The number of cells and genes are stored and automatically updated while
using `dropbead`'s functions

```{r}
length(mms@cells) # number of cells in the sample
length(mms@genes) # number of genes detected
```

Some of the barcodes might correspond to the same cells (essentially after
using the `DetectBeadSynthesisErrors` Drop-seq tool) and have to be collapsed.
Such cells are identified by

```{r}
listCellsToCollapse(mms)
```

In the example above there are 5 pairs of cells which need to be collapsed.
This is straightforward with using

```{r}
mms <- collapseCellsByBarcode(mms)
```

We can verify that the number of cells is now the correct one

```{r}
length(mms@cells)
```

## Handling mixed species samples

Calling `classifyCellsAndDoublets` separates the species and returns a `data.frame` with 
number of transcripts per cell per species. The `threshold` controls the purity of the 
resulting cells, while `min.trans` is the minimum number of UMIs required to keep a cell.

```{r}
head(classifyCellsAndDoublets(mms, 
                              threshold = 0.9, 
                              min.trans = 1000))
```

The output of `classifyCellsAndDoublets` can be directly send for plotting

```{r, fig.width=10, fig.height=7.1}
plotCellTypes(classifyCellsAndDoublets(mms, threshold = 0.9, min.trans = 5000))
```

The number of genes per cell is computed via 

```{r}
head(computeGenesPerCell(mms))
```

and similarly for transcripts (UMIs) with the `computeTranscriptsPerCell` function.
Their output can be send directly for plotting and visualization

```{r}
plotViolin(computeTranscriptsPerCell(mms), 
           attribute = "UMIs")
plotHistogram(computeTranscriptsPerCell(mms), 
              attribute = 'UMIs')
```

The above functions are polymorphic and can be also used for `SingleSpeciesSample`
objects. Splitting the samples is always performed by internal functions automatically. 
If the user wants to restrict to only one species, this is done by the
`splitMixedSpeciesSampleToSingleSpecies` function. which returns a list of the two
`SingleSpeciesSample` objects.

```{r}
# Extracting the human cells as a separate sample for further analysis
h <- splitMixedSpeciesSampleToSingleSpecies(mms,
                                            threshold = 0.9)[[1]]
class(h)
```

# Single species

## Filtering functions
There are a couple of functions to remove low quality cells and genes, such as

```{r}
h.f1 <- keepBestCells(h, num.cells = 100) # keep only the top 100 cells
h.f2 <- keepBestCells(h, min.num.trans = 1000) # keep cells with at least 1000 UMIs
h.f3 <- removeLowQualityCells(h, min.genes = 2000) # remove cells which don't express at least 2000 genes
h.f4 <- removeLowQualityGenes(h, min.cells = 3) # remove genes which are not expressed in at least 3 cells
```

with obvious usage. 

## Comparing gene expression measurements
Reproducibility and correlations of different samples are easily assessed
via the `compareGeneExpressionLevels` function, for instance,

```{r}
compareGeneExpressionLevels(h.f2, h.f1, 
                            name1 = 'Drop-seq with >= 1000 UMIs per cell',
                            name2 = 'Drop-seq only 100 cells',
                            method = 'pearson')
```

If bulk data is available then its correlation pwith the Drop-seq sample is
directly assessed with the function `compareSingleCellsAgainstBulk`. Note that
bulk has to be in a 1-column `data.frame` format with genes as the `rownames`
and RPKM values (as default, raw counts are also accepted)

```{r}
compareSingleCellsAgainstBulk(h.f1, 
                              log2(h.f2@dge[, 1, drop=F]+1))
```

## Further exploration
It is instructive to assess the mitochondrial content of the single cell
samples. This is done as below and the percentages can be sent for plotting

```{r, warning = FALSE}
head(computeMitochondrialPercentage(h), 5)
plotMitochondrialContent(list(100-computeMitochondrialPercentage(h.f1),
                              100-computeMitochondrialPercentage(h.f3)),
                         log_scale = FALSE)
```

`Dropbead` offers also an implementation of the algorithm described in 
Macosko et. al. 2015 for classification of the cell cycle phases. The following
R-packages `xlsx`, `rJava` and `xlsxjars` are required for this

```{r}
phases <- assignCellCyclePhases(h)
```
